Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-05 application.tex
Comments on Section 7 for suggested edits.  Hopefully they are coherent, but ask if not.

\section{Application to protoplanetary disks}\label{application} 
\emph{I suggest the following edits
\begin{itemize}
\item Instead of starting over again with the full energy equation focus on giving a phyiscal values for $\Lambda$ and $t_c$ already introduced.
\item Don't describe what we don't do, especially at the outset.  Say what's done, and then if essential list caveats.
\end{itemize}

@@

\subsection{Radiative diffusion}
\emph{the following sentence is too long and has a dangling "this".}

@@

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin.
\emph{the disk isn't (necessarily) opt. thin, the pert. is}
 In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
fluctuations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the PPD model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv 
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(r, z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\ 
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].         
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, 
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}
It is instructive to compare the above thermal timescale 
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for opacity scales
$\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$, at $z=0$ and $z=3H$. We
also plot Eq. \ref{bcrit_mmsn} in each panel.

Consider first the MMSN shown in the middle panels of 
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, is useful in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height, as assumed in our analysis. We thus infer that the VSI can
operate at tens of AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere (see also Fig. \ref{bcrit_mmsn2d}). 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in the next section.

Lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}, shifts the curves inwards but also raises
them. In fact, lowering the opacity by an order of magnitude renders  
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is 
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

@@

@@

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale given by
Eq. \ref{beta_mmsn_simp}.  Fig. \ref{beta_compare} show examples of
the thermal timescales at $r=5$AU and $r=50$AU for a
perturbation wavenumber $\khat=30$. We also plot the 
corresponding critical thermal timescales evaluated for the MMSN
(Eq. \ref{bcrit_mmsn}).  
 \begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$, which are 
boundary-dependent.  
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and 
Disturbances with large $\khat$ are also more likely subject to viscous 
decay, as discussed above.

>>>>

\section{Application to protoplanetary disks}\label{application} 
\emph{For this section. There are 3 main ideas, and I think they should be the subsections, more comments related to this are in the text itself:
\begin{itemize}
\item Developing the cooling model in opt thin and this regimes, including MMSN value
\item Comparison to expected $\beta_{crit}$
\item Linear growth with MMSN cooling law
\end{itemize}}

\emph{Suggested edits for this intro
\begin{itemize}
\item Instead of starting over again with the full energy equation focus on giving a phyiscal values for $\Lambda$ and $t_c$ already introduced.
\item Don't describe what we don't do, especially at the outset.  Say what's done, and then if essential list caveats.
\end{itemize}}

@@

\subsection{Radiative diffusion}
\emph{I think 7.1-7.4 should be combined to a single section on the "Cooling Model".  As that clearly is a single idea and osme of the subsections are too trivial. Conversion to subsubsections as needed. However the application to beta-crit could be split off as 7.2.  Otherwise that material which has a crucial figure discussion is a bit buried.}

\emph{the following sentence is too long and has a dangling "this".}

@@

\subsection{Newtonian cooling}\label{newton_cool}
\emph{There are no rules against subsections this short, but I think it's bad style.}

For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin.\\
\emph{Above, the disk isn't (necessarily) opt. thin, the pert. is} \\
 In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
fluctuations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
\begin{align}
  t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}
We note that $t_\mathrm{cool}$ does not explicitly depend on $\rho$.

@@

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the PPD model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv 
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(r, z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\ 
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].         
\end{align} 
We refer to the case with $\gamma=1.4$, $\mu=2.33$ and the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN) and $\beta=\beta_\mathrm{MMSN}$, \\
\emph{I would keep $\kappa$ scaling in this eqn, since it's used in plots.  Then just say  that $\kappa_d = 1$ is fiducial case. }
\begin{align}\label{beta_mmsn_simp}
  \beta_\mathrm{MMSN} = 3.9\times10^{-3} r_\mathrm{AU}^{9/14}\left( 1 +
    1.9\times10^7\hat{\rho}^2r_\mathrm{AU}^{-33/7}\khat^{-2}\right). 
\end{align}
\emph{I would explain here that when $k < c\rho \kappa^m r^n$ (with correct values for $c, m, n$) cooling is optically thin, as this is useful for explaining figure.}
% where
% \begin{align}
%   \
% %  \frac{3.7\times10^{-3}}{\mu\left(\gamma-1\right)\hat{\kappa}_d\hat{T}^5}, 
% \end{align}
% and
% \begin{align}
%   \khat_0 = 4.4\times10^{3}\hat{\kappa}_d\hat{T}^2\hat{\Sigma}.
% \end{align}

\emph{Sub-section break  ( 7.2) for the "$\beta_{crit}$ in the MMSN" as this is break from the derivation (and evaluation) of the cooling time (which is fairly general) to the application to the VSI.}

It is instructive (\emph{advice: don't say that "it is instructive", just be instructive}) to compare the above thermal timescale 
with the criterion Eq. \ref{iso_vsi_cond} evaluated for the MMSN,     
\begin{align}\label{bcrit_mmsn}
  % \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
  % -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
  \beta_\mathrm{crit,MMSN} = 0.024r_\mathrm{AU}^{2/7}. 
\end{align}

\emph{This paragraph currently merely copies the figure caption, which is a waste and not illuminating to read.  Better to give some description of what the figure is showing. (see below for ideas)}\\
In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ $\gamma=1.4$, $\hat{T}=\hat{\Sigma}=1$ for opacity scales
$\hat{\kappa}_d=0.1, \,1 $ (MMSN) and $10$, at $z=0$ and $z=3H$. \\
\emph{Here you can describe the general features e.g. diffusive and thin cooling regions, which will help later discussions.}\\
We also plot Eq. \ref{bcrit_mmsn} in each panel.  \emph{This sentence explains nothing and is likely to be of little help to the reader.  Instead wait to describe the science, e.g. after describing the general shape of the $\beta$ curves,  you can start the discussion below with "Fig () also compares the MMSN cooling times to $\beta_{crit}$.}

Consider first the MMSN shown in the middle panels of 
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, is useful in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height, as assumed in our analysis. \\
\emph{above sentence too long and confusing.  "useful" is a vague term.  
"where the curves overlap" is also confusing, forcing the reader to look at the figure in the middle of a complex sentence.  Better to have already explained above how the $\beta$ curves show that opt. thin cooling is height indep. (even though should be obvious from equations).  Then the comparison to $\beta_{crit}$ can be made as something like "Since beta-crit was derived for a height-independent cooling time,  the direct comparison our MMSN cooling model is valid only for modes with optically thin cooling. "} 
We thus infer that the VSI can
operate at tens of AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$ 
may be smaller or greater than $\beta_\mathrm{crit}$ depending on
$z$ as the disk transitions from an optically thick midplane to an
optically thin atmosphere (see also Fig. \ref{bcrit_mmsn2d}). 
%Given a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at
%the optically thick midplane, but with increasing height the disk 
%becomes optically thin and $\beta<\beta_\mathrm{crit}$. 
Our analytical criteria $\beta < \beta_\mathrm{crit}$ is not valid for
vertically non-uniform $\beta$, and we must turn to numerical
solutions of the linear problem to determine whether or not the VSI
operates. We do this in the next section.    \\
\emph{a better organized discussion might end on this point.  Might not be easy to rearrange to do this, but it would ideal to have the seamless flow to what you do next.}

Lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}, shifts the curves inwards but also raises
them.\\
\emph{this graphical discussion is not very inuitive (the main problem) and even arbitrary (lowering vs. shifting to the left).  Better to say what's happening physically, i.e. `With higher opacity, optically thin cooling faster and diffusive cooling is slower, explaining how the curves shift.} \\
In fact, lowering the opacity by an order of magnitude renders  
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This is 
expected to stabilize the disk. Raising the opacity, as shown in 
the lower panels of Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal
timescales, but shifts the curves outward, so the instability is
restricted to larger radii.

@@

\emph{the following statement is not generally true, so it should not be stated without caveat}

@@

\emph{The following viscosity discussion might be better placed elsewhere.  Since the issue also comes up for numerical solutions (not just the comparison to beta-crit here), maybe a short concluding subsection to summarize all the arguments on the "vialibity of short wavelength VSI".}

@@

\subsection{VSI in the MMSN}
Here we solve the linear stability problem in the MMSN with the scale and
height dependent thermal relaxation timescale given by
Eq. \ref{beta_mmsn_simp}. 
 Fig. \ref{beta_compare} show examples of
the thermal timescales at $r=5$AU and $r=50$AU for a
perturbation wavenumber $\khat=30$. We also plot the 
corresponding critical thermal timescales evaluated for the MMSN
(Eq. \ref{bcrit_mmsn}).  \\
\emph{For this figure to belong here (and not in the previous subsection, with the other comparisons of $\beta$ and $\beta_{crit}$) the discussion should be be more focused on the linear analysis.  Again it's better to start by discussing what the figure tells us, than repeating the figure caption. E.g.\ you could use Fig 21 to explain when a height varying $\beta$ is required.  I would also explain (especially if it stays in this subsection) what the growth time is for the r = 5 AU case.  Looks to be $\sim 1000$ orbits, which is remakrable given that $\beta < \beta_{\rm crit}$ outside $z \gtrsim  1.6 H$.?  That kind of detail could be illuminating, as it's nominally surprising given previous evidence that behavior near 5 -- 7 $H$ can affect growth.}

 \begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare}
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    and $r=5$AU for $\khat=30$ (solid lines). The
    corresponding horizontal dashed lines are the critical thermal
    relaxation timescales derived in linear theory. 
    \label{beta_compare}}
\end{figure}

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$, which are 
boundary-dependent.  \\
\emph{above sentence too long and complex, could e.g.\ split off that surface modes are neglected and why.}
%However, surface modes are dependent on
%boundary conditions in a vertically isothermal disk \citep{barker15},  
%and 
Disturbances with large $\khat$ are also more likely subject to viscous 
decay, as discussed above.
